Safe Transferable Regions

### 2017-01-13 paper/introduction.tex
type inference minor updates
Edits to the intro.

This paper presents a \emph{safe} and \emph{efficient} region-based approach
to \emph{data transferable} between a set of concurrent actors, motivated by the following concerns.

Consider the example (of  a producer creating a data structure and transferring it to a consumer)
in Fig.~\ref{fig:motivating-eg}.
This represents a \C{SELECT} query operator that receives a stream of input messages,
each associated with a time window $t$, processed by method \texttt{onReceive}.
Each input message contains a list of inputs, each processed by applying a user-defined function to
create a corresponding output.  Multiple messages with the same timestamp
are permitted and messages with different timestamps may arrive out of order.
An invocation of method
\texttt{OnNotify} indicates that no more input messages with a timestamp
$t$ will be subsequently delivered.  At this point, the operator completes
the processing for time window $t$ and sends a corresponding output message
to its successor.

@@

\mypara{Safe Transferable Regions.}
In this paper, we present a language-based approach to cleanly encapsulated
transferable data and a safe and efficient implementation of such data based
on regions. 
%
A region is a block of memory that is allocated
and freed in one shot, often in constant time. A region may contain
one or more contiguous range of memory locations, and individual
objects may be dynamically allocated within the region over time,
while they are deallocated en masse when the region is freed.  Thus, a
region is a good fit for a transferable data-structure.  In
Fig.~\ref{fig:motivating-eg}, the output to be constructed for each
time window $t$ (i.e., \C{map[t]}) can be a separate region that is
allocated when the first message with timestamp $t$ arrives, and
deallocated after \C{map[t]} is transferred in \C{onNotify}.

@@

\mypara{Type System.}
Formally, our type system introduces \emph{region parameters} (for both
classes and methods) and uses constrained parametric polymorphism
over these parameters, where the constraints capture outlives constraints
between the region parameters.

@@

\mypara{Type Inference.}
One of the key contributions of this paper is a sound type inference algorithm
that eliminates the need for users to write region type annotations.
The users write programs in the underlying language that provides
primitives for the users to manipulate regions (create, open, free,
transfer) and to allocate objects in regions, but has no region type
annotations.
% This, we believe, significantly reduces the impediment to adopt our
% approach in practical setting.

Our inference algorithm proceeds in three stages: in the first stage,
we elaborate the program by introducing region parameters (for classes
and methods); in the second stage we generate a set of constraints
that must be satisfied for the program to type check; in the third
stage, we solve the set of constraints, inferring the preconditions of
all classes and methods (in terms of the expected outlives-constraints
between the region parameters).  We show that the algorithm is sound,
and the last stage of constraint solving is complete.
% (using an interprocedural fixed point computation).

\mypara{Evaluation.}
Our work was inspired by~\cite{Broom:HotOS}, which presents evidence that
realistic programs can be written using transferable regions
and that this can yield significant performance improvements.
The system described in~\cite{Broom:HotOS}, however, provides no memory
safety guarantees, which is the subject of our contribution.
We report on an implementation of our type inference algorithm that
was able to identify some unsafe memory accesses and was able to successfully
verify programs correct (once the errors were fixed).

@@

\begin{itemize} 
  \item We present \broom, a \csharp-like typed
  object-oriented language that eschews garbage collection in favour of
  programmer-managed memory regions. \broom{} extends its core language,
  which includes \emph{lambdas} (higher-order functions) and
  \emph{generics} (parametric polymorphism), with constructs to create,
  manage and free static and transferable memory regions. Transferable regions
  are first-class values in \broom.
  % Experiments demonstrate that   \naiad~\cite{naiad} dataflow
  % program using programmer-managed memory regions outperform their
  % GC counterparts by a margin of upto 59\%.

@@

  \item We describe a region type inference algorithm for \broom that
  (a). completely eliminates the need to annotate \broom programs with
  region types, and (b). enables seamless interoperability between
  region-aware \broom programs and legacy standard library code that is
  region-oblivious. The cornerstone of our inference algorithm is a
  novel constraint solver that performs abduction in a partial-order
  constraint domain to infer weakest solutions to recursive
  constraints.

\item We establish soundness of type inference algorithm, and
completeness of constraint solving.

>>>>

% This paper presents a \emph{safe} and \emph{efficient} region-based approach
% to \emph{data transferable} between a set of concurrent actors, motivated by the following concerns.

Computations performed by a network of concurrent communicating actors
often involve data transfer between producers and consumers. Consider
for example the \C{SELECT} query operator shown in
Fig.~\ref{fig:motivating-eg}, which functions as an actor in a
dataflow computation involving a network of other such query
operators. \C{SELECT} receives a stream of input messages, each
associated with a time window $t$, processed by method
\texttt{onReceive}.  Each input message contains a list of inputs,
each processed by applying a user-defined function to create a
corresponding output.  Multiple messages with the same timestamp are
permitted and messages with different timestamps may arrive out of
order.  An invocation of method \texttt{OnNotify} indicates that no
more input messages with a timestamp $t$ will be subsequently
delivered.  At this point, the operator completes the processing for
time window $t$ and sends a corresponding output message to its
successor.

@@

\mypara{Safe Transferable Regions.}
In this paper, we present a language-based approach to cleanly encapsulate
transferable data and a safe and efficient implementation of such data based
on regions. 
%
A region is a block of memory that is allocated
and freed in one shot, often in constant time. A region may contain
one or more contiguous range of memory locations, and individual
objects may be dynamically allocated within the region over time,
while they are deallocated en masse when the region is freed.  Thus, a
region is a good fit for a transferable data-structure.  In
Fig.~\ref{fig:motivating-eg}, the output to be constructed for each
time window $t$ (i.e., \C{map[t]}) can be a separate region that is
allocated when the first message with timestamp $t$ arrives, and
deallocated after \C{map[t]} is transferred in \C{onNotify}.

@@

\mypara{Type System.}
Formally, our type system introduces \emph{region parameters} (for both
classes and methods) and uses constrained parametric polymorphism
over these parameters, where the constraints capture outlives constraints
between the region parameters.
The type system may be seen as a form of ownership type system~\cite{OwnershipSurvey},
with a region being the owner of all objects allocated in that region.

@@

\mypara{Type Inference.}
One of the key contributions of this paper is a type inference
algorithm that eliminates the need for users to write region type
annotations.  The users write programs in the underlying language that
provides primitives for the users to manipulate regions (create, open,
free, transfer) and to allocate objects in regions, but has no region
type annotations.
% This, we believe, significantly reduces the impediment to adopt our
% approach in practical setting.

Our inference algorithm proceeds in three stages: in the first stage,
we elaborate the program by introducing region parameters (for classes
and methods); in the second stage we generate a set of constraints
that must be satisfied for the program to type check; in the third
stage, we solve the set of constraints, inferring the preconditions of
all classes and methods (in terms of the expected outlives-constraints
between the region parameters).  We show that the algorithm is sound,
and the stages of constraint generation and solving are complete
(i.e., the algorithm is complete relative to the first stage of
elaboration).
% (using an interprocedural fixed point computation).

\mypara{Evaluation.}
Our work was inspired by~\cite{Broom:HotOS}, which presents evidence
that realistic programs can be written using transferable regions and
that this can yield significant performance improvements.
While~\cite{Broom:HotOS} addresses the engineering challenges in
extending a managed runtime with transferable regions, it adopts an ad
hoc approach insofar as language design is concerned, exposing the
region functionality through an unsafe API, and in the process losing
the safety guarantees. Our contribution is an alternative approach
that is grounded in sound theory and restores the language safety
guarantees. The utility of our approach is demonstrated by a prototype
implementation of our type inference algorithm that was able to
identify unsafe memory accesses among the benchmarks extracted
from~\cite{Broom:HotOS}.

@@

\begin{itemize} 
  \item We present \broom, a \csharp-like typed object-oriented
  language that supports programmer-managed memory regions. \broom{}
  extends its core language, which includes \emph{lambdas}
  (higher-order functions) and \emph{generics} (parametric
  polymorphism), with constructs to create, manage and free static and
  transferable memory regions. Transferable regions are first-class
  values in \broom.
  % Experiments demonstrate that   \naiad~\cite{naiad} dataflow
  % program using programmer-managed memory regions outperform their
  % GC counterparts by a margin of upto 59\%.

@@

  \item We describe a region type inference algorithm for \broom that
  (a). completely eliminates the need to annotate \broom programs with
  region types, and (b). enables seamless interoperability between
  region-aware \broom programs and legacy standard library code that
  is region-oblivious. Our inference algorithm is based on a novel
  constraint solver that performs abduction in a partial-order
  constraint domain to infer weakest solutions to recursive
  constraints.

\item We establish the soundness and relative completeness of the type
inference algorithm.
